\ch Printing Utilities
This chapter describes the utility programs PRINT and FF.
Both of these programs assume that the printer is an I.D.S.`460G
"Paper Tiger"; they can easily be adapted to other printers.

\se PRINT
\mr sysut
PRINT is used to print files.
It does not alter the line structure of a file by truncating or wrapping
long lines, but it does paginate.
It should not be used for files that are already paginated except with the
"T" option.

\pa Invoke PRINT either by typing \bd PRINT \ed or
\bd PRINT filename parameters \ed
If you use the first form, PRINT will first display the parameters that
it accepts, and will then wait for you to enter a file name followed
by parameters.
If you use the second form, PRINT will print the file immediately
without displaying the options.
\pa If you do not give it any parameters, PRINT will print the file
using 12 characters/inch, 6 lines/inch, and 60 lines/page.
You can use parameters to change the format of the printed file;
for example, the command
\bd PRINT MOUSE.TEX B C16 N3 X2 \ed
would print two copies (X2) of the file MOUSE.TEX in boldface type (B), with 
16 characters/inch (C16), and line numbers of 3 digits (N3).
Here is a complete list of the parameters that PRINT accepts.
A parameter starts with a letter, and some parameters may be followed by
a number or by two numbers.
\sp 2 B \im 20 Boldface characters.  Note that boldface characters are
twice the width of normal characters.  The default is
'off', that is, normal weight characters.
\sm 0 \sp 2 C <num> \im 20 <num> may be 10, 12, or 16; its value determines 
the pitch (characters/inch) of the printing.  The default value is 
12 characters/inch.
\sm 0 \sp 2 E <m>,<n> \im 20 Print lines <m> through <n> inclusive from the 
file. The default is to print the entire file.
\sm 0 \sp 2 G <num> \im 20 Set tab positions every <num> columns.  The default
value for <num> is 8, corresponding to the editor's default tab positions.
\sm 0 \sp 2 L <num> \im 20 Set line spacing to <num>/48 inches.  The default
value is <num> = 8, so the line spacing is 1/6 inches.
\sm 0 \sp 2 M <num> \im 20 Set the left margin to <num> columns.  The default
value is <num> = 0, so printing starts as close to the left as possible.
\sm 0 \sp 2 N <num> \im 20 Print a line number of <num> digits at the start of
each line.  If <num> = 0, the default value, line numbers are not printed.
\sm 0 \sp 2 P <num> \im 20 Print <num> lines on each page.  The default value
is <num> = 60.  If you use L to change the line spacing, use P accordingly.
The parameter P0 has a special effect: it inhibits PRINT from transmitting 
form-feeds at all.
\sm 0 \sp 2 T <m>,<n> \im 20 Print pages <m> through <n> of a file which
already contains form-feed characters; for example, a file created by 
TP. 
\sp 2 In this mode, TP calls the assembly language procedure _copy.
For some reason that I have not figured out yet, _copy ignores the
ETX/ACK protocol of the printer.
Consequently the "T" mode does not work on my system.
Any suggestions?
\sm 0 \sp 2 V \im 20 Print with proportional spacing.
The default is even spacing.
\sm 0 \sp 2 X <num> \im 20 Print <num> copies.  The default is <num> = 1.
\sm 0

\se FF
This command initializes the printer and may also send form feeds to it.
The command
\bd FF <num> \ed
sends <num> form feeds to the printer, where <num> is a small positive
number.
Use the command
\bd FF 2 \ed
to remove a listing from the printer.
\pa FF always sets the printer to 12 characters/inch, 6 lines/inch,
and fixed space non-bold printing.

\ch Pascal Development Utilities

This chapter describes the programs PP and XREF.
They are designed to work together, and they make certain assumptions about
the way in which Pascal programs are developed.

\se PP
PP pretty-prints Pascal programs; it is a descendent of a program written
by Ledgard et al [SIGPLAN Notices, _12, 7, July 1977, 101-5], but its authors
would probably have difficulty recognizing it in its present form.
The command \bd PP PROG \ed reads a Pascal program from the file PROG.PAS and
writes a pretty-printed version of it to the file PROG.PPP.

\pa PP is not idempotent: if you try to pretty-print a file that has
already been pretty-printed, you may find that PP has inserted unnecessary
blanks.
Thus PP assumes the following programming convention: PAS files are
maintained in a compact form with no indentation, several statements
on a line, etc., and the PPP file is used for printing only.

\pa PP changes all Pascal reserved words (BEGIN, END, WHILE, etc.)
to upper case.
XREF (described below) cross-references only symbols that begin with a
lower case letter, and it will therefore not cross-reference reserved words 
in files created by PP.

\pa PP is a fairly "conservative" pretty-printer in that it gives you a
certain amount of control over the final layout of the program.
It does this not by means of embedded directives, but by retaining
certain features of the input file.
For example, PP will always ensure that a line beginning "PROCEDURE"
is preceded by a blank line; if you always put three blank lines before a
procedure declaration, PP will retain these blank lines.
The same applies to spaces embedded in a line.
Although PP usually tries to put each statement on a separate line,
it is not ruthless about this.
If you write "ELSE`IF`..." on one line, PP will not insert a line break
between "ELSE" and "IF", for example.

\pa The best way to understand PP is to look at its controlling loop, a
_WHILE statement at the very end of the program.
PP reads Pascal source and recognizes certain tokens, including some of the
reserved words and some punctuation symbols.
It associates three sets, _selected, _dindsym, and _terminators, with each
special token, and the members of these sets control its actions.
The formatting conventions that it follows can be altered by judiciously
changing entries in the sets.

\pa It is important to note that PP does not parse the program in any sense.
This is both a strength and a weakness.
A strength because PP is able to pretty-print programs with syntax errors,
and it can therefore be used to track down errors such as mis-matched
\nu /
_BEGIN/END pairs.
A weakness because there are certain situations which it cannot easily be made
to recognize: the most important of these is an _ELSE clause in a _CASE
statement.` (Why didn't Intersystems use _OTHERWISE like everybody else?)

\se XREF
XREF is a program that cross-references Pascal programs that have been
pretty-printed.
The command
/bd XREF PROG /ed
causes XREF to read from the file PROG.PPP and write to a new file
called PROG.XRT. 
\pa XREF does not cross-reference symbols inside comments of the form
"{`...`}", symbols inside quoted strings, or symbols that do not start
with a lower case letter.
It does not write a listing of the program being cross-referenced,
but the line numbers that it prints in the cross-reference table are
consistent with those produced by PRINT.
You can use the following submit file, called PP.SUB, to pretty-print 
and cross-reference a Pascal program, print the results, 
and delete the temporary files:
\bd \rb
PP $1
XREF $1
PRINT $1.PPP N4
PRINT $1.XRT
ERA $1.PPP
ERA $1.XRT
\ed \sb
Note that the parameter $1 must not be given an extension.
For example:
\bd SUBMIT PP MYPROG \ed

\pa XREF uses 26 binary trees to store user identifiers.
The first letter of an identifier determines the tree in which it will
be stored.
An entry in the tree consists of the identifier, a pointer to a list of
line numbers, and pointers to left and right subtrees.
Line numbers are pushed onto the list as the program is scanned,
and the list is reversed prior to printing.

\pa The problem thta is most likely to beset XREF is memory space.
It uses the assembly language function _space to check available memory
before making each entry and it will stop scanning and print the report
if the amount of available memory falls below 100 bytes.
This has not actually happened so far; XREF seems to be capable of
cross-referencing Pascal programs of several thousand lines while
running in 58K.
